---
title: "Getting started with pleiodbr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pleiodbr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
db_path <- "/local-scratch/data/pleiodb/main.pleiodb"
db_available <- dir.exists(db_path)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = db_available,
  fig.width = 8,
  fig.height = 4,
  dpi       = 120
)
if (!db_available)
  message("Database not found at ", db_path, " — all chunks skipped.")
```

`pleiodbr` reads `.pleiodb` databases — compact binary archives of GWAS
z-scores across thousands of traits — directly from R.

## Opening a database

```{r open}
library(pleiodbr)
library(ggplot2)
library(dplyr)

db <- open_pleiodb("/local-scratch/data/pleiodb/main.pleiodb")
db
```

The connection holds variant and trait tables in memory and reads z-scores
on demand from the binary chunks.

---

## Parsing ALIDs

`parse_alid()` splits ALID strings into chromosome, position, effect allele,
and other allele — useful for building annotation tables or MR inputs.

```{r parse-alid}
parse_alid(c("16:53800954_C_T", "19:45412079_C_T", "2:21231524_T_C"))
```

---

## PheWAS — one variant across all traits

```{r phewas-run}
# APOE ε4 proxy: near-perfect LD with rs429358
pw <- phewas(db, "19:45412079_C_T")
nrow(pw)
pw
```

### PheWAS plot

```{r phewas-plot, fig.height=4}
pw_ann <- pw |>
  left_join(db$traits[, c("trait_id", "trait_name")], by = "trait_id") |>
  arrange(pval) |>
  mutate(x = row_number(), log10p = -log10(pval), sig = pval < 5e-8)

top_labels <- pw_ann |> filter(pval < 1e-20)

ggplot(pw_ann, aes(x, log10p, colour = sig)) +
  geom_point(size = 0.7, alpha = 0.6) +
  geom_hline(yintercept = -log10(5e-8), linetype = "dashed", colour = "grey40") +
  ggrepel::geom_text_repel(data = top_labels, aes(label = trait_name),
                           size = 2.5, max.overlaps = 15) +
  scale_colour_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
                      guide = "none") +
  labs(x = "Trait (ranked by p-value)", y = expression(-log[10](p)),
       title = "PheWAS: 19:45412079_C_T (APOE region)") +
  theme_bw(base_size = 11)
```

---

## GWAS — all variants for one trait

```{r gwas-run}
bmi <- gwas(db, "ukb-b-19953")
nrow(bmi)
```

### Manhattan plot

`manhattan_plot()` handles the chromosome offsets, alternating colours, and
optional imputed-variant highlighting automatically.

```{r manhattan, fig.height=4}
manhattan_plot(bmi, title = "Body mass index (BMI) — N ≈ 461,460")
```

---

## Top hits

```{r tophits}
hits <- tophits(
  db,
  traits = c("ukb-b-19953", "ebi-a-GCST006867"),
  pval   = 5e-8
)
hits |>
  left_join(db$traits[, c("trait_id","trait_name")], by = "trait_id") |>
  count(trait_name, wt = NULL)
```

---

## Associations — arbitrary variant × trait block

```{r assoc}
ldl_instruments <- c(
  "19:45412079_C_T",  # APOE region
  "1:55014160_C_T",   # PCSK9 region
  "2:21870368_A_G"    # APOB region
)
assoc <- associations(
  db,
  variants = ldl_instruments,
  traits   = c("ebi-a-GCST90018961",   # LDL cholesterol
               "ebi-a-GCST003116")     # Coronary artery disease
)
assoc
```

---

## Mendelian randomisation: LDL → CAD

`to_twosamplemr()` converts any pleiodbr result tibble directly into the
format expected by `TwoSampleMR`.

, "ebi-a-GCST90013864"

```{r mr-prep}
exposure <- tophits(db, "ukb-b-19953") |>
  to_twosamplemr("exposure", trait_name = "BMI")

outcome <- associations(db, exposure$SNP, "ebi-a-GCST90013864") |>
  to_twosamplemr("outcome", trait_name = "Coronary artery disease")

dat <- TwoSampleMR::harmonise_data(exposure, outcome, action = 1)
res <- TwoSampleMR::mr(dat)
res[, c("method", "nsnp", "b", "se", "pval")]
```

### MR scatter plot

```{r mr-scatter, fig.width=5.5, fig.height=5}
TwoSampleMR::mr_scatter_plot(res, dat)
```

---

## Phenotypic correlation (rho)

```{r rho-heatmap, fig.width=5.5, fig.height=4.5}
library(tidyr)

traits_5 <- c(
  "ukb-b-19953",        # BMI
  "ebi-a-GCST006867",   # Type 2 diabetes
  "ebi-a-GCST90018961", # LDL cholesterol
  "ebi-a-GCST003116",   # Coronary artery disease
  "ebi-a-GCST90002412"  # LDL (larger N)
)

rho_long <- rho(db, traits_5, traits_5)

labels <- db$traits |>
  filter(trait_id %in% traits_5) |>
  mutate(label = paste0(sub(" \\(.*", "", trait_name), "\n(", trait_id, ")")) |>
  select(trait_id, label)

rho_plot <- rho_long |>
  left_join(labels, by = c("trait_id_1" = "trait_id")) |> rename(l1 = label) |>
  left_join(labels, by = c("trait_id_2" = "trait_id")) |> rename(l2 = label)

ggplot(rho_plot, aes(l1, l2, fill = rho)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(rho, 2)), size = 3) +
  scale_fill_gradient2(low = "#4575B4", mid = "white", high = "#D73027",
                       midpoint = 0, limits = c(-1, 1), name = "ρ") +
  labs(x = NULL, y = NULL, title = "Phenotypic correlations") +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
---

## Regional locus plot

```{r regional, fig.height=3.5}
fto <- phewas(db, "16:53e6-54e6") |>
  filter(trait_id == "ukb-b-19953") |>
  mutate(pos_mb = as.numeric(sub(".*:(\\d+)_.*", "\\1", variant_id)) / 1e6)

ggplot(fto, aes(pos_mb, -log10(pval), colour = imputed, shape = imputed)) +
  geom_point(size = 1.8, alpha = 0.85) +
  scale_colour_manual(values = c("FALSE" = "steelblue", "TRUE" = "#E07B39"),
                      labels = c("Observed", "Imputed")) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17),
                     labels = c("Observed", "Imputed")) +
  geom_hline(yintercept = -log10(5e-8), linetype = "dashed", colour = "grey40") +
  labs(x = "Position on chr16 (Mb)", y = expression(-log[10](p)),
       colour = NULL, shape = NULL,
       title = "FTO locus — BMI (chr16:53–54 Mb)") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
```
