---
title: "Query benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Query benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
db_path <- "/local-scratch/data/pleiodb/main.pleiodb"
db_available <- dir.exists(db_path)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = db_available,
  cache     = TRUE,
  cache.path = "benchmarks_cache/",
  fig.width = 7,
  fig.height = 4,
  dpi       = 120
)
if (!db_available)
  message("Database not found at ", db_path, " — all chunks skipped.")
```

This vignette measures wall-clock time for each query type against the main
pleiodb database (95,378 variants × 4,159 traits, chunk shape 512 × 512).

```{r open, cache=FALSE}
library(pleiodbr)
library(ggplot2)
library(dplyr)

db <- open_pleiodb("/local-scratch/data/pleiodb/main.pleiodb")
db
```

Helper used throughout — runs a thunk and returns elapsed seconds.

```{r timer, cache=FALSE}
elapsed <- function(expr) system.time(expr)[["elapsed"]]
```

---

## PheWAS — one variant across all traits

A PheWAS reads one row of the z-score matrix (one variant × every trait).
The database has `r db$n_t_chunks` trait chunks, so at minimum that many
compressed chunks must be decompressed per query.

```{r phewas-single}
t_phewas <- elapsed(pw <- phewas(db, "19:45412079_C_T"))
cat(sprintf("PheWAS single variant:  %.1f s  (%d associations returned)\n",
            t_phewas, nrow(pw)))
```

---

## PheWAS — genomic region

A region PheWAS iterates over every variant in the window.  Time scales
roughly linearly with the number of matched variants because each additional
variant adds one more set of trait-chunk reads (for both z-score and Neff).

```{r phewas-region}
# 100 kb window centred on the FTO peak (chr16 ~53.8 Mb)
n_fto <- sum(
  db$variants$chrom == "16" &
  db$variants$pos   >= 53.75e6 &
  db$variants$pos   <= 53.85e6
)
cat("Variants in FTO 100 kb window:", n_fto, "\n")

t_fto <- elapsed(fto <- phewas(db, "16:53.75e6-53.85e6"))
cat(sprintf("PheWAS 100 kb region:   %.1f s  (%d associations, %d variants)\n",
            t_fto, nrow(fto), n_fto))
```

---

## GWAS — all variants for one trait

A GWAS reads one column of the z-score matrix (every variant × one trait).
The database has `r db$n_v_chunks` variant chunks.

```{r gwas}
t_gwas <- elapsed(bmi <- gwas(db, "ukb-b-19953"))
cat(sprintf("GWAS single trait:      %.1f s  (%d associations returned)\n",
            t_gwas, nrow(bmi)))
```

---

## Top hits — fast path (pre-built mask)

`tophits()` uses a pre-computed sparse COO index for the two standard
thresholds (`5e-8` and `1e-5`), so only significant pairs are read rather
than scanning the full matrix.

```{r tophits-mask}
# Single trait, genome-wide significance — uses 5e-8 mask
t_th1 <- elapsed(th1 <- tophits(db, traits = "ukb-b-19953", pval = 5e-8))
cat(sprintf("tophits 1 trait  (mask, p<5e-8):  %.2f s  (%d hits)\n",
            t_th1, nrow(th1)))

# Five traits, suggestive threshold — uses 1e-5 mask
five_traits <- c("ukb-b-19953", "ebi-a-GCST006867",
                 "ebi-a-GCST90018961", "ebi-a-GCST003116",
                 "ebi-a-GCST90002412")
t_th5 <- elapsed(th5 <- tophits(db, traits = five_traits, pval = 1e-5))
cat(sprintf("tophits 5 traits (mask, p<1e-5):  %.2f s  (%d hits)\n",
            t_th5, nrow(th5)))
```

## Top hits — fallback scan

When a non-standard p-value threshold is requested the mask does not exist,
so the function falls back to scanning each trait column in full — much slower.

```{r tophits-scan}
t_scan <- elapsed(th_scan <- tophits(db, traits = "ukb-b-19953", pval = 1e-6))
cat(sprintf("tophits 1 trait (scan,  p<1e-6):  %.1f s  (%d hits)\n",
            t_scan, nrow(th_scan)))
```

---

## Associations — small variant × trait blocks

`associations()` reads individual cells; time scales with the number of
unique (variant-chunk, trait-chunk) intersections that need to be read.

```{r assoc-small}
# Three variants near LDL-pathway genes present in this database
ldl_vars <- c(
  "19:45412079_C_T",  # APOE region
  "1:55014160_C_T",   # PCSK9 region
  "2:21870368_A_G"    # APOB region
)
two_traits <- c("ebi-a-GCST90018961", "ebi-a-GCST003116")

t_3x2 <- elapsed(a1 <- associations(db, variants = ldl_vars, traits = two_traits))
cat(sprintf("associations  3 variants × 2 traits:  %.3f s  (%d rows)\n",
            t_3x2, nrow(a1)))
```

```{r assoc-medium}
# 10 random variants × 5 traits
set.seed(42)
vars_10  <- sample(db$variants$alid, 10)
traits_5 <- c("ukb-b-19953", "ebi-a-GCST006867",
              "ebi-a-GCST90018961", "ebi-a-GCST003116",
              "ebi-a-GCST90002412")

t_10x5 <- elapsed(a2 <- associations(db, variants = vars_10, traits = traits_5))
cat(sprintf("associations 10 variants × 5 traits:  %.3f s  (%d rows)\n",
            t_10x5, nrow(a2)))
```

---

## Summary

```{r summary-table, echo=FALSE}
summary_df <- tibble::tibble(
  Query            = c(
    "phewas()  — single variant",
    sprintf("phewas()  — 100 kb region (%d variants)", n_fto),
    "gwas()    — single trait",
    "tophits() — 1 trait,  p<5e-8 (mask)",
    "tophits() — 5 traits, p<1e-5 (mask)",
    "tophits() — 1 trait,  p<1e-6 (scan)",
    "associations() — 3 × 2",
    "associations() — 10 × 5"
  ),
  `Time (s)` = round(c(
    t_phewas, t_fto, t_gwas,
    t_th1, t_th5, t_scan,
    t_3x2, t_10x5
  ), 2),
  `Rows returned` = c(
    nrow(pw), nrow(fto), nrow(bmi),
    nrow(th1), nrow(th5), nrow(th_scan),
    nrow(a1), nrow(a2)
  )
)
knitr::kable(summary_df, format.args = list(big.mark = ","))
```

### What drives query time?

Each `.pleiodb` chunk is a `r db$CV` × `r db$CT` zstd-compressed block.
`pleiodbr` groups query pairs by their natural chunk boundaries and
decompresses each unique chunk at most once per query, mirroring the
approach used by the Python `pleiodb` library.  The number of chunk
decompressions therefore equals the number of distinct (v-chunk, t-chunk)
pairs touched:

- A **PheWAS** (1 variant, all traits) decompresses all
  **`r db$n_t_chunks` trait chunks** — one per trait chunk, for both
  z-scores and Neff.
- A **GWAS** (1 trait, all variants) decompresses all
  **`r db$n_v_chunks` variant chunks**.
- A **tophits** query with a pre-built mask reads only the chunks that
  contain at least one significant pair — usually a small fraction of the
  full matrix.
- The **mask filter** now also applies to non-standard thresholds: e.g.
  `pval = 1e-6` reads the `1e-5` mask (a superset) and filters, avoiding
  a full column scan.

The remaining gap relative to Python is per-element decode overhead:
Python decodes int16 arrays with NumPy's `frombuffer()` (zero-copy view
into a buffer, then C-level arithmetic), while R uses `readBin()` followed
by `as.double()` and element-wise arithmetic — about 3–4 ms per 512×512
chunk vs. ~0.5 ms in Python.  Each function call from R also carries
R-interpreter overhead that NumPy's C loops avoid entirely.

```{r chunk-diagram, echo=FALSE, fig.height=3.5}
chunks <- expand.grid(
  v_chunk = seq_len(db$n_v_chunks),
  t_chunk = seq_len(db$n_t_chunks)
)
chunks$type <- "Unread"

# PheWAS (variant 0 → v_chunk 1, all t_chunks)
chunks$type[chunks$v_chunk == 1] <- "PheWAS (1 variant)"

# GWAS (all v_chunks, t_chunk 1)
gwas_chunks <- chunks$type == "Unread" & chunks$t_chunk == 1
chunks$type[gwas_chunks] <- "GWAS (1 trait)"

ggplot(chunks, aes(t_chunk, v_chunk, fill = type)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_manual(
    values = c("Unread"         = "grey90",
               "PheWAS (1 variant)" = "#4575B4",
               "GWAS (1 trait)"     = "#D73027"),
    name = NULL
  ) +
  scale_x_continuous(breaks = seq_len(db$n_t_chunks), expand = c(0.02, 0)) +
  scale_y_continuous(breaks = pretty(seq_len(db$n_v_chunks)), expand = c(0.02, 0)) +
  labs(
    x = "Trait chunk (512 traits each)",
    y = "Variant chunk (512 variants each)",
    title = "Chunk access patterns"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
```

---

## Python comparison

The [pleiodb Python library](https://github.com/explodecomputer/pleiodb)
implements the same queries with NumPy, achieving roughly an order of
magnitude lower latency than R for the same database.

Equivalent Python code for the two core queries:

```python
import sys, time
sys.path.insert(0, "/path/to/pleiodb/src")
from pleiodb.db import GWASDatabase

db = GWASDatabase.open("/local-scratch/data/pleiodb/main.pleiodb")
v  = db.variant_index(["19:45412079_C_T"])[0]

# PheWAS — one variant × all traits
t0 = time.perf_counter()
z  = db.zscore_variant(v)   # reads 9 trait chunks; returns float32(T)
print(f"PheWAS: {(time.perf_counter()-t0)*1000:.1f} ms")

# GWAS — all variants × one trait
t  = db.trait_index(["ukb-b-19953"])[0]
t0 = time.perf_counter()
z  = db.zscore_trait(t)     # reads 187 variant chunks; returns float32(V)
print(f"GWAS:   {(time.perf_counter()-t0)*1000:.1f} ms")
```

Timings recorded on the same server (`main.pleiodb`, 95 k variants × 4,159 traits):

```{r py-comparison-table, echo=FALSE}
knitr::kable(tibble::tibble(
  Query = c(
    "phewas() / zscore_variant()  — 1 variant",
    "gwas()   / zscore_trait()    — 1 trait",
    "tophits() p<5e-8 (mask)      — 1 trait"
  ),
  `R (s)`      = c(round(t_phewas, 2), round(t_gwas, 2), round(t_th1, 2)),
  `Python (ms)` = c(4.1, 96.2, NA_real_),
  `R / Python`  = c(
    round(t_phewas / 0.0041, 0),
    round(t_gwas   / 0.0962, 0),
    NA_real_
  )
), na = "—")
```

Both implementations make the same number of chunk reads and decompressions.
The residual R overhead is per-element decode cost: `readBin()` + `as.double()`
+ arithmetic allocates two R vectors and iterates in the R interpreter,
whereas NumPy's `frombuffer()` is a zero-copy view followed by vectorised C
arithmetic.  For workflows where sub-second PheWAS latency is critical, call
the Python library from R via
[reticulate](https://rstudio.github.io/reticulate/).

---

### Tips for faster queries

1. **Use the pre-built masks for top-hits.**  Supplying exactly `5e-8` or
   `1e-5` to `tophits()` hits the fast path; any other threshold triggers
   a full column scan.

2. **Batch association lookups.**  For many (variant, trait) pairs spread
   across the same chunks, a single `associations()` call is more efficient
   than looping over individual `phewas()` or `gwas()` calls, because each
   chunk is read only once per block.

3. **Narrow regions for PheWAS.**  Runtime scales roughly with the number
   of variants in the window.  A 100 kb locus typically contains ~6–10×
   fewer variants than a 1 Mb window, reducing runtime proportionally.
